Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-08-15 paper2.qmd
Psychological methods, submission

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two key issues in the CJ literature. First, it critiques the reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce (BTL) model in the analysis of CJ data. Specifically, it argues that while the assumptions of equal discriminal dispersions and zero correlation simplify the trait measurement model, they may fail to capture the complexity of some traits or stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between trait measurement and statistical inference in CJ applications. It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.
  To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods. This extension integrates CJ's core theoretical principles alongside key assessment design features. It then translates these elements into a statistical model for analyzing dichotomous CJ data.
  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions. It also encourages researchers across the social sciences to adopt more robust and interpretable methodologies.
date: last-modified
bibliography: references.bib
---

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, empirical evidence of judges’ biases remains scarce in the CJ literature [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. This gap likely persists not from a lack of interest or research but because the identification of such biases often depends on ad-hoc detection methods, like 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022]. To overcome this limitation, the present study treated judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli becomes an observable outcome only through judges' perceptions, which may exhibit bias.

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian inference
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two key issues in the CJ literature. First, it critiques the reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce (BTL) model in the analysis of CJ data. Specifically, it argues that while the assumptions of equal discriminal dispersions and zero correlation simplify the trait measurement model, they may fail to capture the complexity of some traits or stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between trait measurement and statistical inference in CJ applications. It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.

  To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods. This extension integrates CJ's core theoretical principles alongside key assessment design features. It then translates these elements into a statistical model for analyzing dichotomous CJ data.

  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions. It also encourages researchers across the social sciences to adopt more robust and interpretable methodologies.
date: last-modified
bibliography: references.bib
---

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, empirical evidence of judges' biases remains scarce in the CJ literature [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. This gap likely persists not from a lack of interest or research but because the identification of such biases often depends on ad-hoc detection methods, like 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022]. To overcome this limitation, the present study treated judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli becomes an observable outcome only through judges' perceptions, which may exhibit bias.
